---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# qgcomp.multi

<!-- badges: start -->
<!-- badges: end -->

## Overview

`qgcomp.multi` is an extension of quantile g-computation (Keil et al., 2020) to settings with multiple exposure mixtures and their interaction.

In many environmental and epidemiologic studies, exposures arise in distinct groups (e.g., metals, phthalates, phenols) that may jointly influence health outcomes. While standard quantile g-computation estimates the effect of a single mixture, `qgcomp.multi` allows users to:

* Estimate the effect of two exposure mixtures simultaneously
* Evaluate interactions between mixtures
* Fit the model using either quantized exposures or the original exposures scales via `q = NULL`
* Obtain interpretable effect estimates via marginal structural models (MSMs)


## Installation

You can install the development version from GitHub:

``` r
devtools::install_github("cmowrer13/qgcomp.multi")
```

## Basic Usage

Fit a two-mixture quantile g-computation model:

```{r example, eval = FALSE}
library(qgcomp.multi)

fit <- qgcomp.glm.multi.boot(
  f = Y ~ X1 + X2 + X3 + W1 + W2 + W3 + C,
  data = dataset,
  mix1 = c("X1", "X2", "X3"),
  mix2 = c("W1", "W2", "W3"),
  interaction = TRUE,
  q = 4,
  B = 100
)

fit$coef_table
```

## Simulated Example

Generate a dataset with known mixture effects:

```{r}
library(qgcomp.multi)

dat <- sim_mixture_data(
  n = 1000,
  pA = 3,
  pB = 3,
  rho_within_A = 0.3,
  rho_within_B = 0.3,
  rho_between = 0.2,
  psi1 = 0.5,
  psi2 = 0.3,
  psi12 = 0.2,
  seed = 123
)

fit <- qgcomp.glm.multi.boot(
  f = Y ~ X1 + X2 + X3 + W1 + W2 + W3 + C,
  data = dat,
  mix1 = c("X1", "X2", "X3"),
  mix2 = c("W1", "W2", "W3"),
  B = 100
)

fit$coef_table
```

## Computational Considerations

The g-computation step requires prediction over a grid of intervention levels for both mixtures, which can become computationally intensive.

To address this, `qgcomp.multi` includes a Monte Carlo approximation:

* `MCsize` = nrow(data) → full computation
* `MCsize` < nrow(data) → approximate marginalization

This can substantially reduce runtime in large datasets while preserving model structure.

## Package Structure

Core functions include:

* `qgcomp.glm.multi.boot()` — main estimation function
* `msm_fit()` — internal MSM estimation routine
* `quantize_mixtures()` — exposure quantization
* `sim_mixture_data()` — simulation data generator

## References

Keil AP, Buckley JP, O’Brien KM, Ferguson KK, Zhao S, White AJ. A Quantile-Based g-Computation Approach to Addressing the Effects of Exposure Mixtures. Environmental Health Perspectives. 2020;128(4):047004.

## Development

This package is under active development. Contributions, issues, and suggestions are welcome:

https://github.com/cmowrer13/qgcomp.multi
