---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# qgcomp.multi

<!-- badges: start -->
<!-- badges: end -->

## Overview

`qgcomp.multi` extends quantile g-computation (Keil et al., 2020) to settings
with two exposure mixtures and their interaction.

In many environmental and epidemiologic studies, exposures come in distinct
groups, such as metals, phthalates, or phenols, and investigators may want to
estimate the effect of shifting one mixture while allowing the effect of that
shift to depend on the level of another mixture. `qgcomp.multi` is built for
that setting.

The package currently supports:

* estimation with exactly two mixtures, supplied through `mix1` and `mix2`
* optional interaction between the two mixture intervention variables
* quantized analyses with `q >= 2`
* original-scale analyses with `q = NULL`
* native multiple-imputation pooling through `qgcomp.glm.multi.mi()`
* optional bootstrap-level parallel execution for ordinary fits and native MI workflows
* broom-style `tidy()` and `glance()` methods for single-fit and pooled MI objects
* MSM-based prediction, plotting, diagnostics, confidence regions, and sensitivity helpers
* estimand-scale choices for additive and ratio-scale MSM summaries

The package works by fitting an outcome regression, computing predicted
counterfactual means under joint interventions on the two mixtures, and then
fitting a marginal structural model (MSM) to summarize that intervention-response
surface.

For causal interpretation, the usual identifying conditions for g-computation
still matter: causal consistency, conditional exchangeability, positivity for the
interventions under study, and adequate specification of the outcome model.
The package does not test or guarantee those assumptions.

## Installation

You can install the development version from GitHub:

``` r
devtools::install_github("cmowrer13/qgcomp.multi")
```

## Basic Usage

Fit a two-mixture quantile g-computation model:

```{r example, eval = FALSE}
library(qgcomp.multi)

fit <- qgcomp.glm.multi(
  f = Y ~ X1 + X2 + X3 + W1 + W2 + W3 + C,
  data = dataset,
  mix1 = c("X1", "X2", "X3"),
  mix2 = c("W1", "W2", "W3"),
  interaction = TRUE,
  q = 4,
  B = 100,
  seed = 13
)

fit
summary(fit)
coef(fit)
vcov(fit)
confint(fit)
formula(fit)
nobs(fit)
```

The returned object is a fitted `qgcompmulti` model object. In routine use,
the recommended workflow is:

* fit the model with `qgcomp.glm.multi()`
* inspect the fitted object with `print()` or by typing `fit`
* call `summary(fit)` for a fuller report
* use standard generics such as `coef()`, `vcov()`, `confint()`, `formula()`,
  and `nobs()` to extract specific fitted-model components
* optionally supply `seed` when you want bootstrap and Monte Carlo draws to be
  reproducible across repeated fits
  
The public interface also includes:

* `predict()` for MSM-based prediction, direct contrasts, and exact prediction
  targets
* `plot()` for fitted MSM surface displays
* `support()`, `diagnostics()`, and `adequacy()` for model checking
* `mcsize_sensitivity()`, `b_sensitivity()`,  and `q_sensitivity()` for practical 
  robustness checks
* `qgcomp.glm.multi.mi()` for native Rubin-pooled multiple-imputation analysis
* optional `parallel` and `workers` controls for bootstrap-level parallelism
* `tidy()` and `glance()` methods for single-fit and pooled MI reporting
* `confregion()` for bootstrap-covariance chi-squared confidence regions
  
## Specifying the Outcome Model

The `f` argument is the outcome regression formula. In most applications it
should include:

* the outcome variable
* all exposure variables named in `mix1` and `mix2`
* covariates needed for confounding control
* any nonlinear terms or transformations you want in the outcome model

The `mix1` and `mix2` arguments define which exposure variables belong to each
mixture. They do not automatically add variables to the formula, so the
exposures still need to appear in `f`.

One implementation detail is worth stating explicitly. If `interaction = TRUE`,
the package fits:

* an outcome model that includes a product between the sums of the
  mixture components, and
* an MSM that includes the `psi1 * psi2` interaction term

So `interaction = TRUE` does not just change the summary model at the end. It
also changes the nuisance outcome regression used in the g-computation step.
  
## Interpreting the MSM coefficients

The main estimands are the coefficients of the marginal structural model (MSM):

* `psi1` is the Mixture 1 main effect
* `psi2` is the Mixture 2 main effect
* `psi1:psi2` is the mixture interaction when `interaction = TRUE`

When `q` is an integer, the MSM is indexed by quantized intervention levels.
In that setting, a one-unit change in `psi1` or `psi2` corresponds to a
simultaneous one-quantile increase in all components of the corresponding
mixture.

When `q = NULL`, the interpretation changes. The package does not use quantile
scores. Instead, it builds a `3 x 3` fit-time intervention grid from pooled
25th, 50th, and 75th percentile values within each mixture, and under a given
intervention every component in a mixture is set to the same pooled
mixture-specific value on the analysis scale. In that setting, the MSM
coefficients are defined with respect to that original-scale intervention coding.

The fitted scale depends on the outcome family and the requested
`estimand_scale`. Gaussian fits use mean differences. Binomial-logit fits now
default to odds ratios, with risk differences available by setting
`estimand_scale = "risk_difference"`. Poisson-log fits now default to rate
ratios, with additive expected-count summaries available through
`estimand_scale = "mean_difference"`.

For odds-ratio and rate-ratio estimands, the stored MSM coefficients and
standard errors stay on the log fitting scale. Print, summary, `confint()`, and
`tidy()` add display-scale quantities where those are easier to report. MSM
predictions and plots remain on the response scale unless a direct MSM contrast
explicitly requests `contrast_scale = "estimand"`.

The coefficients should be read as summaries of the intervention-response
surface implied by the fitted outcome model. When that surface is close to
linear in the chosen intervention coding, the MSM coefficients are especially
easy to interpret. When the surface is more curved, the same coefficients are
still defined, but they are doing more approximation work.

## Simulated Example

Generate a dataset with known mixture effects and inspect the fitted model
through the model-object workflow:

```{r}
library(qgcomp.multi)

dat <- sim_mixture_data(
  n = 1000,
  pA = 3,
  pB = 3,
  rho_within_A = 0.3,
  rho_within_B = 0.3,
  rho_between = 0.2,
  psi1 = 0.5,
  psi2 = 0.3,
  psi12 = 0.2,
  seed = 123
)

fit <- qgcomp.glm.multi(
  f = Y ~ X1 + X2 + X3 + W1 + W2 + W3 + C,
  data = dat,
  mix1 = c("X1", "X2", "X3"),
  mix2 = c("W1", "W2", "W3"),
  B = 100,
  seed = 13
)

fit

coef(fit)

confint(fit)
confregion(fit, parm = c("psi1", "psi2"))
```

For binary and count outcomes, read the coefficient table together with the
active estimand scale. The effect-scales vignette gives the details, including
why ratio coefficients are fit on a log scale while predictions stay on the
response scale.

For a fuller report that still keeps the MSM as the primary focus, call:

```{r}
summary(fit)
```

## Prediction Workflow

The default prediction target is the fitted marginal structural model (MSM)
surface:

```r
pred_msm <- predict(fit)
head(pred_msm$estimates)
```

You can also request direct MSM contrasts between two intervention regimes:

```r
predict(
  fit,
  type = "msm_contrast",
  from = c(psi1 = 0, psi2 = 0),
  to = c(psi1 = 3, psi2 = 3),
  interval = TRUE
)
```

### MSM versus exact predictions

`qgcomp.multi` distinguishes between:

* MSM-based predictions, which come from the fitted marginal structural model
* exact counterfactual predictions, which come directly from the fitted outcome
  model

The stored fit-time exact surface can be extracted without any additional data:

```r
predict(fit, type = "exact")
```

For arbitrary exact prediction at a new intervention regime, the user must
supply `data`:

```r
predict(
  fit,
  type = "exact",
  data = dat,
  at = c(psi1 = 1, psi2 = 2)
)
```

Supplying `data` is required because an exact counterfactual mean is defined by
averaging predicted potential outcomes over a concrete covariate distribution.
Without user-supplied data, the package can only return the exact surface that
was already computed and stored at fit time.

For MSM-based prediction, user inputs such as `grid`, `at`, `from`, and `to`
are interpreted on the MSM coding scale. That point matters most when
`q = NULL` and `centering = "median"`, because the stored heatmaps are labeled
on the intervention-value scale even though prediction inputs live on the
centered MSM scale. In that setting, `(0, 0)` corresponds to the pooled median
intervention for both mixtures.

Interval support is limited to MSM-based predictions and direct MSM contrasts.
Use `method = "percentile"` or `method = "basic"` for those bootstrap intervals.

## Plotting Workflow

The default plot is a heatmap of the fitted MSM surface over the stored
intervention grid:

```{r}
plot(fit)
```

You can switch to a contour display:

```{r}
plot(fit, style = "contour")
```

To visualize uncertainty, use the slice-based interval display:

```{r, eval = FALSE}
plot(
  fit,
  interval = TRUE,
  slice = list(var = "psi2", value = 1)
)
```

When `q = NULL`, the stored heatmap axes are labeled with the pooled
25th/50th/75th percentile intervention values used at fit time, rather than
the centered MSM coordinates used internally when `centering = "median"`.

## Diagnostics Workflow

Support, bootstrap behavior, and MSM adequacy are all available through public
diagnostic helpers:

```r
support(fit)
diagnostics(fit, type = "bootstrap")
adequacy(fit)
```

These diagnostics are kept outside the main model summary so that users can
inspect model behavior directly rather than relying on a crowded `summary()`
output.

Two interpretation notes matter here:

* support diagnostics summarize the intervention grid used to define the stored
  surface; they are not a full proof of causal positivity
* adequacy diagnostics compare the exact fit-time counterfactual surface to the
  fitted MSM surface on the response scale; they are checks on MSM
  approximation, not tests of whether the outcome model is true

Adequacy is evaluated on the stored fit-time grid. So a good adequacy result
means that the MSM tracks the exact fit-time surface well on that grid. It does
not prove that the surface is globally linear between those points or beyond
them.
  
## Sensitivity Workflow

The package includes dedicated helpers for sensitivity to Monte Carlo size, the
number of bootstrap iterations, and quantization choice:

```{r, eval = FALSE}
mcsize_sensitivity(
  f = Y ~ X1 + X2 + X3 + W1 + W2 + W3 + C,
  data = dat,
  mix1 = c("X1", "X2", "X3"),
  mix2 = c("W1", "W2", "W3"),
  MCsize_values = c(250, 500, 1000),
  q = 4,
  B = 100,
  seed = 13
)

b_sensitivity(
  f = Y ~ X1 + X2 + X3 + W1 + W2 + W3 + C,
  data = dat,
  mix1 = c("X1", "X2", "X3"),
  mix2 = c("W1", "W2", "W3"),
  B_values = c(50, 100, 200),
  q = 4,
  MCsize = nrow(dat),
  seed = 13
)

q_sensitivity(
  f = Y ~ X1 + X2 + X3 + W1 + W2 + W3 + C,
  data = dat,
  mix1 = c("X1", "X2", "X3"),
  mix2 = c("W1", "W2", "W3"),
  q_values = c(3, 4, 5),
  B = 100,
  seed = 13
)
```

For `q` sensitivity, raw coefficient magnitudes should not be interpreted
naively across different `q` values. A larger `q` means that a one-quantile
increase is a smaller intervention step, so smaller effect sizes may be
expected mechanically rather than indicating a weaker overall mixture effect.

## Original-Scale Fitting with `q = NULL`

If you do not want to quantize exposures, set `q = NULL`. In that case, the
outcome model is fit on the original analysis scale and the MSM is built over a
pooled `3 x 3` intervention grid within each mixture.

That intervention needs to be described carefully. For each mixture, the package
computes pooled 25th, 50th, and 75th percentile values across the components in
that mixture. Under a given intervention, every component in the mixture is then
set to the same pooled mixture-specific value.

```r
fit_cont <- qgcomp.glm.multi(
  f = Y ~ X1 + X2 + X3 + W1 + W2 + W3 + C,
  data = dataset,
  mix1 = c("X1", "X2", "X3"),
  mix2 = c("W1", "W2", "W3"),
  interaction = TRUE,
  q = NULL,
  centering = "median",
  B = 100,
  seed = 13
)

summary(fit_cont)
coef(fit_cont)
confint(fit_cont)
```

With `q = NULL`, the MSM coefficients are not one-quantile effects. They are
defined with respect to the original-scale intervention coding used to fit the
MSM.

If `centering = "median"`, the MSM is fit on intervention values centered at
the pooled median within each mixture. That makes the intercept correspond to
the predicted outcome when both mixtures are set to their pooled medians.

This option can be useful when the original analysis scale matters, but users
should still check whether the implied pooled intervention is scientifically
sensible for the exposures at hand.

## Native Multiple Imputation Workflow

The native multiple-imputation wrapper fits and pools completed-data analyses:

* `qgcomp.glm.multi.mi()` fits one `qgcomp.glm.multi()` model per completed
  imputation
* it pools the marginal structural model (MSM) coefficients and covariance
  internally using Rubin's rules
* the returned `qgcompmulti_mi` object is built for inference
* pooled prediction, pooled diagnostics, pooled bootstrap interval methods, and
  pooled confidence regions are not yet implemented

Use this route when you want `qgcomp.multi` to handle the imputation
normalization, repeated fitting, and Rubin pooling steps itself.

```{r mi-example, eval = FALSE}
dat <- sim_mixture_data(
  n = 150,
  pA = 3,
  pB = 3,
  rho_within_A = 0.3,
  rho_within_B = 0.3,
  rho_between = 0.2,
  psi1 = 0.5,
  psi2 = 0.3,
  psi12 = 0.2,
  return_quantized = FALSE,
  seed = 123
)

imp_list <- lapply(
  seq_len(3),
  function(i) {
    dat_i <- dat
    dat_i$X1 <- dat_i$X1 + rnorm(nrow(dat_i), sd = 0.02 * i)
    dat_i$W2 <- dat_i$W2 + rnorm(nrow(dat_i), sd = 0.02 * i)
    dat_i
  }
)

fit_mi <- qgcomp.glm.multi.mi(
  f = Y ~ X1 + X2 + X3 + W1 + W2 + W3 + C,
  data = imp_list,
  mix1 = c("X1", "X2", "X3"),
  mix2 = c("W1", "W2", "W3"),
  interaction = TRUE,
  q = 4,
  B = 10,
  seed = 13
)

summary(fit_mi)
coef(fit_mi)
confint(fit_mi)
```

If you want the stored per-imputation fitted objects as well as the pooled
result, set `keep_fits = TRUE`:

```{r mi-keep-fits, eval = FALSE}
fit_mi_keep <- qgcomp.glm.multi.mi(
  f = Y ~ X1 + X2 + X3 + W1 + W2 + W3 + C,
  data = imp_list,
  mix1 = c("X1", "X2", "X3"),
  mix2 = c("W1", "W2", "W3"),
  interaction = TRUE,
  q = NULL,
  centering = "median",
  B = 10,
  seed = 17,
  keep_fits = TRUE
)

fit_mi_keep$mi_info$keep_fits
fit_mi_keep$fits$imputation_fits[[1]]
```

`mids` objects from the optional `mice` package are also supported:

```{r mi-mids, eval = FALSE}
library(mice)

dat_miss <- dat
dat_miss$X1[sample.int(nrow(dat_miss), 10)] <- NA_real_
dat_miss$W2[sample.int(nrow(dat_miss), 10)] <- NA_real_

mids_obj <- mice(dat_miss, m = 3, maxit = 1, printFlag = FALSE)

fit_mi_mids <- qgcomp.glm.multi.mi(
  f = Y ~ X1 + X2 + X3 + W1 + W2 + W3 + C,
  data = mids_obj,
  mix1 = c("X1", "X2", "X3"),
  mix2 = c("W1", "W2", "W3"),
  interaction = TRUE,
  q = 4,
  B = 10,
  seed = 19
)
```

## Optional Bootstrap-Level Parallelism

The package also supports one level of optional parallelism. The supported
pattern is:

* for ordinary fits, parallelize bootstrap replications within
  `qgcomp.glm.multi()`
* for native MI fits, keep the imputation loop serial and parallelize the
  bootstrap replications inside each per-imputation fit

```{r parallel-example, eval = FALSE}
fit_parallel <- qgcomp.glm.multi(
  f = Y ~ X1 + X2 + X3 + W1 + W2 + W3 + C,
  data = dat,
  mix1 = c("X1", "X2", "X3"),
  mix2 = c("W1", "W2", "W3"),
  interaction = TRUE,
  q = 4,
  B = 20,
  seed = 29,
  parallel = TRUE,
  workers = 2
)

fit_mi_parallel <- qgcomp.glm.multi.mi(
  f = Y ~ X1 + X2 + X3 + W1 + W2 + W3 + C,
  data = imp_list,
  mix1 = c("X1", "X2", "X3"),
  mix2 = c("W1", "W2", "W3"),
  interaction = TRUE,
  q = 4,
  B = 10,
  seed = 31,
  parallel = TRUE,
  workers = 2
)
```

Two behavior notes:

* reproducibility is defined within a fixed backend and execution mode, not as
  exact equality between serial and parallel runs
* `progress = TRUE` is supported only in serial mode and is disabled with a
  clear warning when `parallel = TRUE`
  
## Computational Considerations

The g-computation step requires prediction over a two-dimensional intervention
grid, and the bootstrap repeats that workflow many times. In moderate or large
datasets, that can become expensive.

To address that, `qgcomp.multi` includes a Monte Carlo approximation:

* `MCsize = nrow(data)` means full empirical averaging
* `MCsize < nrow(data)` means approximate marginalization over a random
  subsample

This reduces runtime, but it introduces an additional approximation. In a real
analysis, `MCsize` should be treated as a stability question, not just a tuning
knob.

In practice:

* use a smaller `B` while you are developing code and checking model
  specification
* increase `B` for the final analysis once the model is stable
* keep `MCsize = nrow(data)` when full g-computation is feasible
* consider smaller `MCsize` values only when runtime is a real constraint
* use `mcsize_sensitivity()` to check whether results are stable across
  reasonable `MCsize` choices
* use `seed` when you need bootstrap-based results to be exactly reproducible
  across reruns

## Core Functions and Methods

Core functions include:

* `qgcomp.glm.multi()` — main estimation function
* `qgcompmulti_msm_fit()` — MSM estimation routine
* `quantize_mixtures()` — exposure quantization
* `sim_mixture_data()` — simulation data generator

The standard fitted-model methods include:
 
* `print()` — compact interactive display of the fitted object
* `summary()` — fuller model summary with MSM results in the foreground
* `coef()` — MSM coefficients
* `vcov()` — covariance matrix for the MSM coefficients
* `confint()` — Wald, percentile, and basic confidence intervals for single-fit MSM coefficients
* `confregion()` — fitting-scale confidence regions for selected MSM coefficients
* `formula()` — original fitted formula
* `nobs()` — number of observations actually used

Other public interfaces include:

* `qgcomp.glm.multi.mi()` — native multiple-imputation fitting and pooling
* `predict()` — MSM predictions, contrasts, and exact prediction targets
* `plot()` — MSM surface plotting
* `diagnostics()` — structured model diagnostics
* `support()` — intervention support diagnostic
* `adequacy()` — MSM adequacy diagnostic
* `mcsize_sensitivity()` — repeated-fit sensitivity to Monte Carlo size
* `b_sensitivity()` — repeated-fit sensitivity to number of bootstrap iterations
* `q_sensitivity()` — repeated-fit sensitivity to quantization choice
* `tidy()` and `glance()` — broom-style reporting helpers for single-fit and
  pooled MI objects

## References

Keil AP, Buckley JP, O’Brien KM, Ferguson KK, Zhao S, White AJ. A Quantile-Based g-Computation Approach to Addressing the Effects of Exposure Mixtures. Environmental Health Perspectives. 2020;128(4):047004.

## Development

This package is under active development. Contributions, issues, and suggestions are welcome:

https://github.com/cmowrer13/qgcomp.multi
