---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# qgcomp.multi

<!-- badges: start -->
<!-- badges: end -->

## Overview

`qgcomp.multi` is an extension of quantile g-computation (Keil et al., 2020) to settings with multiple exposure mixtures and their interaction.

In many environmental and epidemiologic studies, exposures arise in distinct groups (e.g., metals, phthalates, phenols) that may jointly influence health outcomes. While standard quantile g-computation estimates the effect of a single mixture, `qgcomp.multi` allows users to:

* Estimate the effect of two exposure mixtures simultaneously
* Evaluate interactions between mixtures
* Fit the model using either quantized exposures or the original exposure scales via `q = NULL`
* Obtain interpretable effect estimates via marginal structural models (MSMs)

The current package interface is designed for analyses with exactly two
mixtures, supplied through `mix1` and `mix2`.

## Installation

You can install the development version from GitHub:

``` r
devtools::install_github("cmowrer13/qgcomp.multi")
```

## Basic Usage

Fit a two-mixture quantile g-computation model:

```{r example, eval = FALSE}
library(qgcomp.multi)

fit <- qgcomp.glm.multi(
  f = Y ~ X1 + X2 + X3 + W1 + W2 + W3 + C,
  data = dataset,
  mix1 = c("X1", "X2", "X3"),
  mix2 = c("W1", "W2", "W3"),
  interaction = TRUE,
  q = 4,
  B = 100,
  seed = 13
)

fit
summary(fit)
coef(fit)
vcov(fit)
confint(fit)
formula(fit)
nobs(fit)
```

The returned object is a fitted `qgcompmulti` model object. In routine use,
the recommended workflow is:

* fit the model with `qgcomp.glm.multi()`
* inspect the fitted object with `print()` or by typing `fit`
* call `summary(fit)` for a fuller report
* use standard generics such as `coef()`, `vcov()`, `confint()`, `formula()`,
  and `nobs()` to extract specific fitted-model components
* optionally supply `seed` when you want bootstrap and Monte Carlo draws to be
  reproducible across repeated fits
  
The `0.3.0` interface then extends into:

* `predict()` for MSM-based prediction, direct contrasts, and exact prediction
  targets
* `plot()` for fitted MSM surface displays
* `support()`, `diagnostics()`, and `adequacy()` for model checking
* `mcsize_sensitivity()` and `q_sensitivity()` for practical robustness checks
  
## Specifying the Outcome Model

The `f` argument is the outcome regression formula. In most applications it
should include:

* the outcome variable
* all exposure variables named in `mix1` and `mix2`
* covariates you want included in the outcome model

The `mix1` and `mix2` arguments define which exposure variables belong to each
mixture. They do not automatically add variables to the formula, so the
exposures should still appear in `f`.
  
## Interpreting the MSM coefficients

The main estimands are the coefficients of the marginal structural model
(MSM):
 
* `psi1` is the Mixture 1 main effect
* `psi2` is the Mixture 2 main effect
* `psi1:psi2` is the mixture interaction when `interaction = TRUE`

When `q` is an integer, these coefficients describe the change in the expected
outcome associated with a one-quantile simultaneous increase in all components
of the corresponding mixture. When `q = NULL`, the coefficients are defined on
the original-scale intervention structure used to build the MSM, so their units
depend on the exposure scales. In both cases, the MSM is fit to predicted
potential outcomes on the response scale.

Interpretation also depends on the outcome type used in the outcome model:
 
* for continuous outcomes, the MSM coefficients are mean differences
* for binary outcomes fit with `family = binomial()`, they are risk differences
* for count outcomes fit with `family = poisson()`, they are differences in expected counts

## Simulated Example

Generate a dataset with known mixture effects and inspect the fitted model
through the model-object workflow:

```{r}
library(qgcomp.multi)

dat <- sim_mixture_data(
  n = 1000,
  pA = 3,
  pB = 3,
  rho_within_A = 0.3,
  rho_within_B = 0.3,
  rho_between = 0.2,
  psi1 = 0.5,
  psi2 = 0.3,
  psi12 = 0.2,
  seed = 123
)

fit <- qgcomp.glm.multi(
  f = Y ~ X1 + X2 + X3 + W1 + W2 + W3 + C,
  data = dat,
  mix1 = c("X1", "X2", "X3"),
  mix2 = c("W1", "W2", "W3"),
  B = 100,
  seed = 13
)

fit

coef(fit)

confint(fit)
```

For a fuller report that still keeps the MSM as the primary focus, call:

```{r}
summary(fit)
```

## Prediction Workflow

The default prediction target is the fitted marginal structural model (MSM)
surface:

```{r}
pred_msm <- predict(fit)
head(pred_msm$estimates)
```

You can also request direct MSM contrasts between two intervention regimes:

```{r}
predict(
  fit,
  type = "msm_contrast",
  from = c(psi1 = 0, psi2 = 0),
  to = c(psi1 = 3, psi2 = 3),
  interval = TRUE
)
```

### MSM versus exact predictions

`qgcomp.multi` distinguishes between:

* MSM-based predictions, which come from the fitted marginal structural model
* exact counterfactual predictions, which come directly from the fitted outcome model
The stored fit-time exact surface can be extracted without any additional data:

```{r}
predict(fit, type = "exact")
```

For arbitrary exact prediction at a new intervention regime, the user must
supply `data`:

```{r}
predict(
  fit,
  type = "exact",
  data = dat,
  at = c(psi1 = 1, psi2 = 2)
)
```

Supplying `data` is required here because an exact counterfactual mean is defined by
averaging predicted potential outcomes over a concrete covariate distribution.
Without user-supplied data, the package does not know which covariate
distribution to average over, so it can only return the exact surface that was
already computed and stored at fit time.

In Version `0.3.0`, interval support is currently limited to MSM-based
predictions.

## Plotting Workflow

The default plot is a heatmap of the fitted MSM surface over the stored
intervention grid:

```{r}
plot(fit)
```

You can switch to a contour display:

```{r}
plot(fit, style = "contour")
```

To visualize uncertainty, use the slice-based interval display:

```{r, eval = FALSE}
plot(
  fit,
  interval = TRUE,
  slice = list(var = "psi2", value = 1)
)
```

When `q = NULL`, the stored heatmap axes are labeled with the pooled
25th/50th/75th percentile intervention values used at fit time, rather than
the centered MSM coordinates used internally when `centering = "median"`.

## Diagnostics Workflow

Support, bootstrap behavior, and MSM adequacy are all available through public
diagnostic helpers:

```{r}
support(fit)
diagnostics(fit, type = "bootstrap")
adequacy(fit)
```

These diagnostics are deliberately kept outside the main model summary so that
users can inspect model behavior explicitly rather than relying on a very dense
`summary()` output.

Two important interpretation notes:
* support diagnostics summarize the intervention grid and are especially
  informative for `q = NULL`
* adequacy diagnostics compare the exact fit-time counterfactual surface to the
  fitted MSM surface on the response scale
  
## Sensitivity Workflow

The package includes dedicated helpers for sensitivity to Monte Carlo size and
quantization choice:
```{r, eval = FALSE}
mcsize_sensitivity(
  f = Y ~ X1 + X2 + X3 + W1 + W2 + W3 + C,
  data = dat,
  mix1 = c("X1", "X2", "X3"),
  mix2 = c("W1", "W2", "W3"),
  MCsize_values = c(250, 500, 1000),
  q = 4,
  B = 100,
  seed = 13
)

q_sensitivity(
  f = Y ~ X1 + X2 + X3 + W1 + W2 + W3 + C,
  data = dat,
  mix1 = c("X1", "X2", "X3"),
  mix2 = c("W1", "W2", "W3"),
  q_values = c(3, 4, 5),
  B = 100,
  seed = 13
)
```

For `q` sensitivity, raw coefficient magnitudes should not be interpreted
naively across different `q` values. A larger `q` means that a one-quantile
increase is a smaller intervention step, so smaller effect sizes may be
expected mechanically rather than indicating a weaker overall mixture effect.

## Original-Scale Fitting with `q = NULL`

If you do not want to quantize exposures, set `q = NULL`. In that case, the
outcome model is fit on the original exposure scales and the MSM is built over
a common pooled intervention grid within each mixture.

```{r example-original-scale, eval = FALSE}
fit_cont <- qgcomp.glm.multi(
  f = Y ~ X1 + X2 + X3 + W1 + W2 + W3 + C,
  data = dataset,
  mix1 = c("X1", "X2", "X3"),
  mix2 = c("W1", "W2", "W3"),
  interaction = TRUE,
  q = NULL,
  centering = "median",
  B = 100,
  seed = 13
)
 
summary(fit_cont)
coef(fit_cont)
confint(fit_cont)
```

With `q = NULL`, the MSM coefficients are interpreted with respect to the
original-scale exposure coding rather than a one-quantile increase.
Choosing `centering = "median"` makes the intercept correspond to the predicted
outcome when both mixtures are set to their pooled medians.

## Computational Considerations

The g-computation step requires prediction over a grid of intervention levels for both mixtures, which can become computationally intensive.

To address this, `qgcomp.multi` includes a Monte Carlo approximation:

* `MCsize` = nrow(data) → full computation
* `MCsize` < nrow(data) → approximate marginalization

This can substantially reduce runtime in large datasets while preserving model structure.

In practice:
 
* use a smaller `B` while developing code and checking model specification
* increase `B` for your final analysis once the model is stable
* keep `MCsize = nrow(data)` when full g-computation is feasible
* consider a smaller `MCsize` when runtime is prohibitive, especially in large samples
* use `seed` when you need the bootstrap-based inferential results to be
  exactly reproducible across reruns

## Core Functions and Methods

Core functions include:

* `qgcomp.glm.multi()` — main estimation function
* `qgcompmulti_msm_fit()` — MSM estimation routine
* `quantize_mixtures()` — exposure quantization
* `sim_mixture_data()` — simulation data generator

The standard fitted-model methods include:
 
* `print()` — compact interactive display of the fitted object
* `summary()` — fuller model summary with MSM results in the foreground
* `coef()` — MSM coefficients
* `vcov()` — covariance matrix for the MSM coefficients
* `confint()` — Wald confidence intervals for the MSM coefficients
* `formula()` — original fitted formula
* `nobs()` — number of observations actually used

New `0.3.0` public methods include:

* `predict()` — MSM predictions, contrasts, and exact prediction targets
* `plot()` — MSM surface plotting
* `diagnostics()` — structured model diagnostics
* `support()` — intervention support diagnostic
* `adequacy()` — MSM adequacy diagnostic
* `mcsize_sensitivity()` — repeated-fit sensitivity to Monte Carlo size
* `q_sensitivity()` — repeated-fit sensitivity to quantization choice

## References

Keil AP, Buckley JP, O’Brien KM, Ferguson KK, Zhao S, White AJ. A Quantile-Based g-Computation Approach to Addressing the Effects of Exposure Mixtures. Environmental Health Perspectives. 2020;128(4):047004.

## Development

This package is under active development. Contributions, issues, and suggestions are welcome:

https://github.com/cmowrer13/qgcomp.multi
