---
title: "Applied Workflow for Two-Mixture Quantile g-Computation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Applied Workflow for Two-Mixture Quantile g-Computation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

# Introduction

`qgcomp.multi` extends quantile g-computation to settings with **two exposure
mixtures** and their interaction.

This document is meant to read like a real analysis walkthrough, not just a list
of function calls. The example uses a simulated environmental health dataset
with:

- Mixture 1: urinary phthalate metabolites
- Mixture 2: urinary phenol biomarkers
- Outcome: a continuous inflammation score

The package can now do more than it could in the early proof-of-concept stage.
The fitted object behaves like an S3 model object, and the package now supports
prediction, plotting, and several diagnostics. This draft focuses on how an
applied analyst would actually use those features.

Two distinctions matter throughout:

- the **outcome model**, which is the regression model you fit to the observed data
- the **marginal structural model (MSM)**, which is the lower-dimensional summary
  that the package fits to predicted counterfactual means

Those are not the same object, and a lot of the interpretation hinges on that.

# What This Vignette Covers

This workflow shows:

- a standard quantized fit
- an original-scale fit with `q = NULL`
- why log transformation and median centering might help with biomarker data
- model extraction with standard S3 generics
- MSM-based prediction and direct contrasts
- exact fit-time surface extraction
- plotting
- intervention support, bootstrap, and adequacy diagnostics
- sensitivity to `MCsize`
- sensitivity to `q`

# Before You Fit Anything

For an applied analysis, it helps to be explicit about the target before you
start.

If you fit a quantized model with `q = 4`, the package constructs interventions
that move every component in a mixture up by one quantile category at the same
time. The MSM coefficient for `psi1` or `psi2` is then a summary of the mean
outcome difference associated with that joint one-quantile shift.

If you fit with `q = NULL`, the package does not work with quantile scores.
Instead, it builds the intervention grid from pooled 25th, 50th, and 75th 
percentile values within each mixture on the analysis scale. In that setting, 
the coefficients are tied to the original-scale intervention coding used to fit 
the MSM.

That choice matters. For biomarker work, it changes both what the model means
and how easy it is to explain.

# Simulate A Biomarker Dataset

We start with a narrative teaching dataset from `sim_biomarker_data()`.

```{r}
library(qgcomp.multi)

dat <- sim_biomarker_data(
  n = 600,
  include_log = TRUE,
  seed = 123
)
```

The raw biomarker concentrations are positive and right-skewed, which is common
for urinary biomarker data. Their scales also differ quite a bit across
chemicals.

```{r}
raw_vars <- c(
  "mep_ng_ml", "mibp_ng_ml", "mehpp_ng_ml",
  "bpa_ng_ml", "bp3_ng_ml", "ppb_ng_ml"
)

round(sapply(dat[raw_vars], median), 2)
```

The dataset also includes:

- `age_years`
- `bmi_kg_m2`
- `smoker`
- `inflammation_score`

Logged biomarker variables are included as well:

```{r}
head(dat[c("ln_mep", "ln_bpa", "inflammation_score")])
```

# Define The Mixtures

For the rest of the vignette, we will use:

```{r}
mix1_raw <- c("mep_ng_ml", "mibp_ng_ml", "mehpp_ng_ml")
mix2_raw <- c("bpa_ng_ml", "bp3_ng_ml", "ppb_ng_ml")

mix1_log <- c("ln_mep", "ln_mibp", "ln_mehpp")
mix2_log <- c("ln_bpa", "ln_bp3", "ln_ppb")
```

# A Note About Example Settings

Some settings in the code below are smaller than what should be used in a final
analysis:

- `B = 20` is used to keep the vignette runnable
- some sensitivity examples use even smaller `B`

That is fine for a teaching document, but not for a paper-ready analysis. In a
real application, you would usually increase `B` enough that bootstrap intervals
and standard errors are stable, and you would check that your conclusions do not
change meaningfully when `MCsize` is increased.

# Standard Quantized Analysis

We start with a standard quantized fit using the **raw biomarker concentrations**.

When `q` is an integer, the exposures are quantized by rank. A monotone
transformation such as `log()` preserves the ordering, so a simple logged-versus-
raw change will usually not alter the quantized intervention coding.

That is why users do not need to log-transform first just to run a quantized
analysis.

```{r}
fit_q <- qgcomp.glm.multi(
  f = inflammation_score ~
    mep_ng_ml + mibp_ng_ml + mehpp_ng_ml +
    bpa_ng_ml + bp3_ng_ml + ppb_ng_ml +
    age_years + bmi_kg_m2 + smoker,
  data = dat,
  mix1 = mix1_raw,
  mix2 = mix2_raw,
  interaction = TRUE,
  q = 4,
  B = 20,
  seed = 13
)
```

Typing the fitted object gives a compact summary:

```{r}
fit_q
```

`summary()` gives a fuller report:

```{r}
summary(fit_q)
```

# Core Model Extractors

The fitted object behaves like a model object, so standard extractors are the
cleanest way to pull out specific components.

```{r}
coef(fit_q)
```

```{r}
vcov(fit_q)
```

```{r}
confint(fit_q)
```

```{r}
formula(fit_q)
nobs(fit_q)
```

# Interpreting The MSM Coefficients

The main estimands are the coefficients of the marginal structural model:

- `psi1`: Mixture 1 main effect
- `psi2`: Mixture 2 main effect
- `psi1:psi2`: interaction between the two mixtures

For the quantized fit, a one-unit change in the mixture intervention level means
a **simultaneous one-quantile increase** in all components of that mixture.

So here:

- `psi1` is the expected change in the outcome for a one-quantile simultaneous
  increase in the phthalate mixture, when the phenol mixture is held fixed at
  the reference intervention level used by the MSM (the lowest quantile)
- `psi2` is the analogous effect for the phenol mixture
- `psi1:psi2` tells us how the effect of one mixture changes as the intervention
  level of the other mixture changes

Because the model includes interaction, the main effects are part of a surface.
They should not be read as one constant effect that applies everywhere.

There is another point worth stating plainly: these are **MSM coefficients**.
They summarize the intervention-response surface implied by the fitted outcome
model. If that surface is close to linear in the intervention coding, the MSM
coefficients are easy to read. If the surface is more curved, the same
coefficients are still defined, but they are doing more approximation work.

# MSM Predictions And Contrasts

The default prediction target is the fitted MSM surface over the stored
intervention grid.

```{r}
pred_q <- predict(fit_q)
head(pred_q$estimates)
```

You can also request a single MSM point prediction:

```{r}
predict(
  fit_q,
  type = "msm_point",
  at = c(psi1 = 1, psi2 = 2),
  interval = TRUE
)
```

And you can compare two intervention regimes directly:

```{r}
predict(
  fit_q,
  type = "msm_contrast",
  from = c(psi1 = 0, psi2 = 0),
  to = c(psi1 = 3, psi2 = 3),
  interval = TRUE
)
```

In a paper, a contrast like that is often more useful than quoting `psi1` and
`psi2` alone, because it maps onto a specific intervention comparison.

# Exact Versus MSM Predictions

`qgcomp.multi` supports two related but different prediction targets:

- **MSM predictions**, which come from the fitted low-dimensional marginal
  structural model
- **Exact predictions**, which come directly from the fitted outcome model under
  the specified intervention regimes

The stored fit-time exact surface can be extracted directly:

```{r}
exact_q <- predict(fit_q, type = "exact")
head(exact_q$estimates)
```

That object contains the exact fit-time counterfactual means that were already
computed when the model was fitted.

If you want the exact mean at a regime that is already on the stored grid, you
can simply subset the stored surface:

```{r}
subset(
  exact_q$estimates,
  intervention_psi1 == 1 & intervention_psi2 == 2
)
```

You can also inspect the stored surfaces together:

```{r}
surf_q <- fitted_surface(fit_q)
names(surf_q)
```

## Why arbitrary exact predictions require user-supplied data

A genuinely new exact prediction is not just a model plug-in value. It is an
average of predicted potential outcomes over some covariate distribution.

That is why arbitrary exact prediction requests need user-supplied `data`.
Without that, the package can only return the fit-time exact surface that was
already stored.

For example, in a quantized fit with `q = 4`, the stored grid contains the
intervention values `0`, `1`, `2`, and `3`. A request like `psi1 = 1.5`,
`psi2 = 2.5` is outside that stored grid and would require a new exact
calculation:

```{r, eval = FALSE}
predict(
  fit_q,
  type = "exact",
  data = dat,
  at = c(psi1 = 1.5, psi2 = 2.5)
)
```

For quantized analyses, interpolation like this is usually not the main goal.
It becomes more natural when `q = NULL`, because users may want predictions at
custom intervention values on the original analysis scale.

In Version `0.3.0`, bootstrap percentile-based interval support is attached to 
MSM-based predictions, but not to arbitrary exact predictions.

# Plot The Fitted MSM Surface

The default plot is a heatmap of the fitted MSM surface.

```{r}
plot(fit_q)
```

A contour plot is also available:

```{r}
plot(fit_q, style = "contour")
```

For uncertainty, the first public plotting form is a slice-based interval
display:

```{r}
plot(
  fit_q,
  interval = TRUE,
  slice = list(var = "psi2", value = 1),
  level = 0.95
)
```

This is often a more readable uncertainty display than trying to put intervals
directly on a full two-dimensional surface.

# Diagnostics For The Quantized Fit

Diagnostics should be part of routine interpretation, not something you check
only when a fit looks odd.

## Intervention Support

```{r}
support(fit_q)
```

For a quantized model with `q = 4`, the support diagnostic is easy to read
because the stored intervention grid is finite and discrete.

Still, this is not a proof of causal positivity. It tells you which
intervention values were used to define the stored surface. It does not prove
that every underlying intervention is equally plausible in the source
population.

## Bootstrap Behavior

```{r}
diagnostics(fit_q, type = "bootstrap")
```

This reports how many bootstrap replications were requested, how many were
retained, and whether any failures were logged.

If failures appear in a real analysis, that is a sign to look more closely at
model specification, numerical stability, and the bootstrap sample size.

## MSM Adequacy

```{r}
adequacy(fit_q)
```

For this basic quantized fit, the adequacy diagnostic may look nearly perfect.
That does **not** mean the package has shown that the outcome model is correct
or that the analysis has recovered the truth.

It answers a narrower question:

> Does the fitted MSM do a good job approximating the intervention-response
> surface implied by the fitted outcome model?

If the fitted outcome model implies a surface that is close to linear in the
chosen intervention coding, the exact fit-time surface and the MSM surface can
line up almost exactly.

One more caution belongs here. The adequacy diagnostic is checked on the stored
fit-time grid. For a quantized fit with `q = 4`, that grid has 16 points. For
`q = NULL`, it has only 9 points. So good adequacy means the MSM works well on
that grid. It does not prove that the exact surface is globally linear between
or beyond those points.

## A more informative adequacy example

To see why adequacy matters, fit a second quantized model with a few nonlinear
terms in the outcome model:

```{r}
fit_q_nl <- qgcomp.glm.multi(
  f = inflammation_score ~
    mep_ng_ml + I(mep_ng_ml^2) +
    mibp_ng_ml + I(mibp_ng_ml^2) +
    mehpp_ng_ml + I(mehpp_ng_ml^2) +
    bpa_ng_ml + bp3_ng_ml + ppb_ng_ml +
    age_years + bmi_kg_m2 + smoker,
  data = dat,
  mix1 = mix1_raw,
  mix2 = mix2_raw,
  interaction = TRUE,
  q = 4,
  B = 20,
  seed = 2024
)
```

```{r}
adequacy(fit_q_nl)
```

This example needs one careful note.

Because `q = 4`, the package quantizes the mixture variables before fitting the
outcome model. So the terms `mep_ng_ml`, `I(mep_ng_ml^2)`, and the rest are operating
on **quantized scores derived from the biomarkers**, not on the original
continuous biomarker values themselves.

That still makes sense as an adequacy demonstration. The point here is to show
that once the outcome model implies a more curved intervention-response surface,
the linear MSM may no longer be an exact summary of that surface.

What does **not** change is the meaning of the intervention step itself. In this
quantized analysis, a one-unit change in the MSM intervention coding is still a
one-quantile simultaneous increase in all components of the mixture.

What changes is how literally you should read the coefficient as a constant
increment. With nonlinear terms in the outcome model, the MSM coefficient is
better thought of as a low-dimensional summary of the exact fit-time surface.

That is exactly what `adequacy()` is checking.

## If you want nonlinearities on the log scale

If your scientific question is about curvature on the continuous logged
biomarker scale itself, the cleaner demonstration is to use `q = NULL` and put
the nonlinear terms in that original-scale outcome model.

The package allows that:

```{r, eval = FALSE}
fit_cont_nl <- qgcomp.glm.multi(
  f = inflammation_score ~
    ln_mep + I(ln_mep^2) +
    ln_mibp + I(ln_mibp^2) +
    ln_mehpp + I(ln_mehpp^2) +
    ln_bpa + ln_bp3 + ln_ppb +
    age_years + bmi_kg_m2 + smoker,
  data = dat,
  mix1 = mix1_log,
  mix2 = mix2_log,
  interaction = TRUE,
  q = NULL,
  centering = "median",
  B = 20,
  seed = 2024
)

adequacy(fit_cont_nl)
```

That is a different teaching example. There, the nonlinear terms really are
nonlinearities in the continuous logged exposures, and the adequacy diagnostic
is asking whether the linear MSM is still a reasonable summary over the stored
original-scale intervention grid.

# Original-Scale Fitting With `q = NULL`

The main reason to use `q = NULL` is that you want the outcome model and the
intervention grid to live on the analysis scale you care about, rather than on
quantile categories.

For biomarker concentrations, that choice matters. Here the logged scale is
easier to work with than the raw concentration scale because:

- the biomarkers are right-skewed
- the raw scales differ substantially across chemicals
- a log scale makes multiplicative differences easier to interpret
- pooled percentile interventions are less awkward on the log scale than on the
  raw scale

So for the `q = NULL` fit, we use the logged biomarkers.

```{r}
fit_cont <- qgcomp.glm.multi(
  f = inflammation_score ~
    ln_mep + ln_mibp + ln_mehpp +
    ln_bpa + ln_bp3 + ln_ppb +
    age_years + bmi_kg_m2 + smoker,
  data = dat,
  mix1 = mix1_log,
  mix2 = mix2_log,
  interaction = TRUE,
  q = NULL,
  centering = "median",
  B = 20,
  seed = 2024
)
```

```{r}
summary(fit_cont)
```

## What the `q = NULL` intervention means

This is where users should to slow down and read carefully.

When `q = NULL`, the package builds a fit-time `3 x 3` intervention grid from
pooled 25th, 50th, and 75th percentile values within each mixture. Under a
given intervention, every component in a mixture is set to the same pooled
mixture-specific value on the analysis scale.

That is not a one-quantile shift. It is an original-scale joint intervention
defined by those common pooled percentile values.

For some applications that is useful. For others it may feel artificial. There
is no way around that judgment call, so it is better to say it directly.

## Why median centering helps here

With `centering = "median"`, the MSM is fit using intervention values centered
at the pooled median within each mixture. The outcome model itself is unchanged;
the centering only changes the MSM coding.

That makes the intercept easier to read. It now corresponds to the predicted
outcome when both mixtures are set to their pooled median logged biomarker
values.

That is usually more interpretable than an intercept tied to some arbitrary
reference value on the raw concentration scale.

Parameter interpretation in this setting is:

- `psi1` is the expected change in outcome for a one-unit increase in the MSM
  intervention coding for mixture 1, on the centered log-scale intervention
  coding used by the fitted MSM
- `psi2` is the analogous one-unit change for mixture 2 on that same MSM scale
- `psi1:psi2` describes how the slope for one mixture changes as the
  intervention level of the other mixture changes

So here, the coefficients are tied to the original-scale intervention structure
used to fit the MSM. They are not one-quantile effects.

# Prediction, Plotting, And Diagnostics For The `q = NULL` Fit

The same downstream workflow works for the original-scale fit.

When `centering = "median"` and you request MSM predictions, the prediction
points are on the reparameterized MSM scale. In practice, that means `(0, 0)`
corresponds to the pooled median intervention for both mixtures.

```{r}
predict(
  fit_cont,
  type = "msm_point",
  at = c(psi1 = 0, psi2 = 0),
  interval = TRUE
)
```

```{r}
plot(fit_cont)
```

For stored `q = NULL` heatmaps, the axis labels are shown on the intervention
value scale rather than on the centered MSM scale. That is a good default,
because those are the values most users will want to see.

```{r}
support(fit_cont)
```

The support diagnostic matters more here than it did in the quantized example,
because the intervention grid now depends directly on the analysis scale and the
pooled percentile construction.

Two cautions are worth keeping in mind:

- if mixture components live on very different scales, a common pooled value may
  be harder to interpret
- good-looking support output still does not mean the intervention is
  scientifically realistic or policy-relevant

In practice, if the intervention grid looks awkward on the raw scale, that is a
good reason to consider transformation, standardization, or a return to a
quantized analysis.

# Sensitivity To Monte Carlo Size

The `mcsize_sensitivity()` helper compares repeated fits while changing only the
Monte Carlo approximation size.

```{r}
mcsize_sensitivity(
  f = inflammation_score ~
    mep_ng_ml + mibp_ng_ml + mehpp_ng_ml +
    bpa_ng_ml + bp3_ng_ml + ppb_ng_ml +
    age_years + bmi_kg_m2 + smoker,
  data = dat,
  mix1 = mix1_raw,
  mix2 = mix2_raw,
  MCsize_values = c(300, 600),
  interaction = TRUE,
  q = 4,
  B = 20,
  seed = 2024,
  keep_fits = FALSE
)
```

The point is not to discover one universally correct `MCsize`, but to
see whether the estimated coefficients, adequacy summaries, and bootstrap
behavior are reasonably stable as the approximation changes.

# Sensitivity To `q`

The `q_sensitivity()` helper holds the rest of the analysis fixed while varying
the number of quantization levels.

```{r}
q_sensitivity(
  f = inflammation_score ~
    mep_ng_ml + mibp_ng_ml + mehpp_ng_ml +
    bpa_ng_ml + bp3_ng_ml + ppb_ng_ml +
    age_years + bmi_kg_m2 + smoker,
  data = dat,
  mix1 = mix1_raw,
  mix2 = mix2_raw,
  q_values = c(3, 4),
  interaction = TRUE,
  B = 8,
  MCsize = nrow(dat),
  seed = 2024,
  keep_fits = FALSE
)
```

The package warns that raw coefficient magnitudes are **not directly comparable across different choices of `q`**.

That warning is important. A larger `q` means a smaller one-quantile step, so
smaller coefficients can appear for purely mechanical reasons even when the
substantive pattern is basically unchanged.

This helper is best treated as a robustness check, not a way to rank effect
sizes across different quantization choices.

# What An Applied User Should Check In A Real Analysis

The software workflow is only part of the job. In a real environmental
epidemiology analysis, I would want to see the analyst think through at least
the following:

- whether the two mixtures were defined a priori or after looking at the data
- whether the outcome model includes the covariates needed for exchangeability
- whether the exposure scale is scientifically sensible for the question at hand
- whether a quantized analysis or an original-scale analysis is easier to defend
- whether the support diagnostic reveals intervention values that are hard to
  interpret
- whether the adequacy diagnostic suggests the MSM is too crude a summary
- whether results are stable to larger `B`, larger `MCsize`, and reasonable
  changes in `q`

# Practical Takeaways

This workflow suggests a workable way to use `qgcomp.multi`:

1. Start with a standard quantized fit when you want a familiar rank-based
   analysis and an intervention that is easy to explain.
2. Use MSM predictions, contrasts, and plots to understand the fitted
   intervention-response surface.
3. Check support, bootstrap behavior, and adequacy before you get too attached
   to a tidy coefficient table.
4. Move to `q = NULL` when the original analysis scale matters enough to justify
   the more complicated intervention definition.
5. For biomarker concentrations, log transformation plus median centering often
   makes original-scale summaries easier to read.
6. Use the sensitivity helpers to look for instability.

