---
title: "Clustered Data and Repeated Measures"
description: >
  Use the id argument to obtain cluster-level bootstrap inference for
  clustered data or repeated measures in qgcomp.multi workflows.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Clustered Data and Repeated Measures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

# Introduction

This article is a short companion to the main workflow vignette. It focuses on
one practical feature that matters whenever rows are not the right resampling
unit for inference:

- clustered sampling, where observations are grouped within higher-level units
- repeated measures, where multiple rows belong to the same individual

In these settings, `qgcomp.multi` can still be used through the usual fitting
interface. The key change is to supply the cluster or subject identifier
through the `id` argument.

When `id` is supplied, the package performs bootstrap resampling at the cluster
level rather than at the individual row level. This changes the uncertainty
calculation to account for within-cluster dependence.


# A small clustered example

The example below creates a small clustered dataset with 50 clusters and 3
observations per cluster. The same setup also works as a simple repeated-
measures template if each cluster ID corresponds to one participant observed
multiple times.

```{r}
library(qgcomp.multi)

n_clusters <- 50
cluster_size <- 3
n_total <- n_clusters * cluster_size

dat <- sim_mixture_data(
  n = n_total,
  pA = 3,
  pB = 3,
  rho_within_A = 0.3,
  rho_within_B = 0.3,
  rho_between = 0.2,
  psi1 = 0.5,
  psi2 = 0.3,
  psi12 = 0.2,
  return_quantized = FALSE,
  seed = 123
)

dat$cluster_id <- rep(seq_len(n_clusters), each = cluster_size)

head(dat[c("cluster_id", "Y", "X1", "W1", "C")])
```

# Fit the model with cluster-level resampling

Supply the cluster identifier through `id`.

```{r}
fit_clustered <- qgcomp.glm.multi(
  f = Y ~ X1 + X2 + X3 + W1 + W2 + W3 + C,
  data = dat,
  mix1 = c("X1", "X2", "X3"),
  mix2 = c("W1", "W2", "W3"),
  interaction = TRUE,
  q = 4,
  B = 5,
  id = "cluster_id",
  MCsize = nrow(dat),
  seed = 13
)
```

This is still the ordinary quantized two-mixture workflow. The difference is
that bootstrap uncertainty is now based on resampling the 50 clusters instead
of resampling the 150 rows one by one.

# Confirm that the fit recognized the clustering

The printed summary reports clustering metadata directly:

```{r}
summary(fit_clustered)
```

You can also inspect the stored fit metadata programmatically:

```{r}
fit_clustered$data_info[c("has_clusters", "cluster_var", "n_clusters", "n_used")]
```

For repeated-measures data, this is the same idea: set `id` to the participant
identifier so that all rows from the same participant move together during the
bootstrap resampling step.

# What the id argument changes

Supplying `id` changes the resampling unit used for bootstrap inference.

That is the right adjustment when:

- observations are clustered within a higher-level unit
- repeated measurements are recorded on the same individual
- rows within a cluster should not be treated as independently resampled units

In practical terms, this lets you keep the same `qgcomp.glm.multi()` fitting
interface while asking the bootstrap to respect the structure of the data.

# What the id argument does not change

The `id` argument is about inference, not about changing the outcome model into
a mixed-effects or subject-specific model.

It does not:

- automatically add random effects
- change the formula interface
- solve every longitudinal modeling question on its own

Instead, it tells `qgcomp.multi` what the correct resampling unit is when the
rows are clustered or repeated within a subject.

# Brief notes on other workflows

The same clustered-resampling idea also works in other package workflows.

- If you fit an original-scale model with `q = NULL`, you can still supply
  `id` in the same way.
- If you use `qgcomp.glm.multi.mi()` with clustered completed datasets, the
  wrapper also supports clustered bootstrap resampling.

For multiply imputed clustered workflows, the completed datasets must share the
same cluster IDs in the same order.

# Practical takeaways

The main question is whether the bootstrap should resample rows or resample
clusters. In `qgcomp.multi`, supplying `id` tells the package to use the
clustered resampling path while keeping the rest of the two-mixture workflow
the same.

# Related Articles

- [Applied Workflow for Two-Mixture Quantile g-Computation](qgcompmulti-workflow.html)
- [Multiple Imputation and Parallel Workflows](qgcompmulti-mi-parallel.html)
- [Diagnostics and Sensitivity Checks](qgcompmulti-diagnostics-sensitivity.html)
