---
title: "Multiple Imputation and Parallel Workflows"
description: >
  Use qgcomp.glm.multi.mi() for multiply imputed analyses and optional
  bootstrap-level parallel execution for ordinary and MI workflows.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multiple Imputation and Parallel Workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

# Introduction

The package includes two workflow-oriented capabilities that often matter
together in practice:

- native multiple-imputation fitting through `qgcomp.glm.multi.mi()`
- optional bootstrap-level parallel execution for ordinary and MI workflows

This article is a short companion to the main workflow vignette. It focuses on
when to use the native MI wrapper, what the pooled object contains, and how the
parallel option fits into larger repeated-fit analyses.

# When to use the native MI wrapper

Use `qgcomp.glm.multi.mi()` when you want `qgcomp.multi` to:

- fit one `qgcomp.glm.multi()` model per imputation
- pool the MSM coefficients and covariance internally using Rubin's rules

The pooled object is focused on inference and summary. It
does not try to reproduce the full prediction and diagnostics surface of a
single-fit `qgcompmulti` object.

# A small completed-data example

```{r}
library(qgcomp.multi)

dat <- sim_mixture_data(
  n = 120,
  pA = 3,
  pB = 3,
  rho_within_A = 0.3,
  rho_within_B = 0.3,
  rho_between = 0.2,
  psi1 = 0.5,
  psi2 = 0.3,
  psi12 = 0.2,
  return_quantized = FALSE,
  seed = 123
)

imp_list <- lapply(
  seq_len(2),
  function(i) {
    dat_i <- dat
    dat_i$X1 <- dat_i$X1 + rnorm(nrow(dat_i), sd = 0.02 * i)
    dat_i$W2 <- dat_i$W2 + rnorm(nrow(dat_i), sd = 0.02 * i)
    dat_i
  }
)
```

Fit the pooled model:

```{r}
fit_mi <- qgcomp.glm.multi.mi(
  f = Y ~ X1 + X2 + X3 + W1 + W2 + W3 + C,
  data = imp_list,
  mix1 = c("X1", "X2", "X3"),
  mix2 = c("W1", "W2", "W3"),
  interaction = TRUE,
  q = 4,
  B = 5,
  seed = 13
)
```

Inspect the pooled result:

```{r}
summary(fit_mi)
coef(fit_mi)
confint(fit_mi)
```

# Retaining per-imputation fits

If you want the full completed-data fits in addition to the pooled result, set
`keep_fits = TRUE`.

```{r, eval = FALSE}
fit_mi_keep <- qgcomp.glm.multi.mi(
  f = Y ~ X1 + X2 + X3 + W1 + W2 + W3 + C,
  data = imp_list,
  mix1 = c("X1", "X2", "X3"),
  mix2 = c("W1", "W2", "W3"),
  interaction = TRUE,
  q = NULL,
  centering = "median",
  B = 5,
  seed = 17,
  keep_fits = TRUE
)

fit_mi_keep$mi_info$keep_fits
fit_mi_keep$fits$imputation_fits[[1]]
```

This is useful when you want pooled inference as the main result but still need
to inspect one or more single-fit `qgcompmulti` objects directly.

# `mids` objects are also supported

The wrapper also accepts `mice::mids` objects when the optional `mice` package
is installed.

```{r, eval = FALSE}
library(mice)

dat_miss <- dat
dat_miss$X1[sample.int(nrow(dat_miss), 8)] <- NA_real_
dat_miss$W2[sample.int(nrow(dat_miss), 8)] <- NA_real_

mids_obj <- mice(dat_miss, m = 3, maxit = 1, printFlag = FALSE)

fit_mi_mids <- qgcomp.glm.multi.mi(
  f = Y ~ X1 + X2 + X3 + W1 + W2 + W3 + C,
  data = mids_obj,
  mix1 = c("X1", "X2", "X3"),
  mix2 = c("W1", "W2", "W3"),
  interaction = TRUE,
  q = 4,
  B = 5,
  seed = 19
)
```

# Optional bootstrap-level parallelism

The package supports one level of optional parallelism.

For ordinary fits, bootstrap replications can be dispatched in parallel inside
`qgcomp.glm.multi()`:

```{r, eval = FALSE}
fit_parallel <- qgcomp.glm.multi(
  f = Y ~ X1 + X2 + X3 + W1 + W2 + W3 + C,
  data = dat,
  mix1 = c("X1", "X2", "X3"),
  mix2 = c("W1", "W2", "W3"),
  interaction = TRUE,
  q = 4,
  B = 20,
  seed = 29,
  parallel = TRUE,
  workers = 2
)
```

For native MI fits, the imputation loop stays serial, but the bootstrap work
inside each completed-data fit can still be parallelized:

```{r, eval = FALSE}
fit_mi_parallel <- qgcomp.glm.multi.mi(
  f = Y ~ X1 + X2 + X3 + W1 + W2 + W3 + C,
  data = imp_list,
  mix1 = c("X1", "X2", "X3"),
  mix2 = c("W1", "W2", "W3"),
  interaction = TRUE,
  q = 4,
  B = 10,
  seed = 31,
  parallel = TRUE,
  workers = 2
)
```

# Practical notes

Several practical limits are worth stating plainly:

- the MI loop itself is still serial
- reproducibility is defined within a fixed backend and execution mode, not as
  exact equality between serial and parallel runs
- `progress = TRUE` is supported only in serial mode and is disabled with a
  clear warning when `parallel = TRUE`
- pooled prediction, pooled diagnostics, pooled bootstrap interval methods, and
  pooled confidence regions are not yet the main target of the pooled MI object

In practice, the native MI wrapper is best used when pooled MSM inference is the
main goal and you want the package to manage the repeated fitting and Rubin
pooling directly.

# Related Articles

- [Applied Workflow for Two-Mixture Quantile g-Computation](qgcompmulti-workflow.html)
- [Effect Scales, Intervals, and Prediction Scales](qgcompmulti-effect-scales.html)
- [Clustered Data and Repeated Measures](qgcompmulti-clustered-data.html)
- [Diagnostics and Sensitivity Checks](qgcompmulti-diagnostics-sensitivity.html)
