---
title: "Diagnostics and Sensitivity Checks"
description: >
  Use support(), diagnostics(), adequacy(), and the sensitivity helpers to
  check how stable and interpretable a qgcomp.multi analysis is.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Diagnostics and Sensitivity Checks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

# Introduction

This article is a short companion to the main workflow vignette. It focuses on
the post-fit tools that help you judge whether a `qgcomp.multi` analysis is
stable and interpretable:

- `support()` for the intervention grid behind the stored surface
- `diagnostics()` for bootstrap behavior
- `adequacy()` for how well the fitted MSM summarizes the exact fit-time surface
- sensitivity helpers for Monte Carlo size, bootstrap size, and quantization choice

These are not replacements for scientific judgment. They are structured checks
that help you see where the analysis is easy to defend and where it needs more
thought.

# Fit a small working example

```{r}
library(qgcomp.multi)

dat <- sim_mixture_data(
  n = 300,
  pA = 3,
  pB = 3,
  rho_within_A = 0.3,
  rho_within_B = 0.3,
  rho_between = 0.2,
  psi1 = 0.5,
  psi2 = 0.3,
  psi12 = 0.2,
  return_quantized = FALSE,
  seed = 123
)

fit <- qgcomp.glm.multi(
  f = Y ~ X1 + X2 + X3 + W1 + W2 + W3 + C,
  data = dat,
  mix1 = c("X1", "X2", "X3"),
  mix2 = c("W1", "W2", "W3"),
  interaction = TRUE,
  q = 4,
  B = 10,
  MCsize = nrow(dat),
  seed = 13
)
```

# Intervention support

```{r}
support(fit)
```

The support diagnostic summarizes the intervention grid used to define the
stored surface. It is especially easy to read for quantized models, where the
grid is finite and discrete.

What it does **not** do is prove causal positivity. It tells you what
intervention values are represented in the stored fit-time surface, not whether
every underlying intervention is equally plausible in the source population.

# Bootstrap behavior

```{r}
diagnostics(fit, type = "bootstrap")
```

This reports how many bootstrap replications were requested, how many were
retained, and whether any failures were logged.

If failures appear in a real analysis, that is a sign to look more closely at
model specification, numerical stability, and covariate sparsity.

# MSM adequacy

```{r}
adequacy(fit)
```

The adequacy diagnostic asks a narrower question than model truth:

> Does the fitted MSM do a good job approximating the intervention-response
> surface implied by the fitted outcome model?

Good adequacy means the MSM tracks the stored exact fit-time surface well on the
fit-time grid. It does not prove that the outcome model is correct, and it does
not prove that the exact surface is globally linear between or beyond those
grid points.

# Sensitivity helpers

The package includes dedicated helpers for practical repeated-fit checks.

For Monte Carlo size:

```{r, eval = FALSE}
mcsize_sensitivity(
  f = Y ~ X1 + X2 + X3 + W1 + W2 + W3 + C,
  data = dat,
  mix1 = c("X1", "X2", "X3"),
  mix2 = c("W1", "W2", "W3"),
  MCsize_values = c(150, 300),
  interaction = TRUE,
  q = 4,
  B = 10,
  seed = 13
)
```

For bootstrap size:

```{r, eval = FALSE}
b_sensitivity(
  f = Y ~ X1 + X2 + X3 + W1 + W2 + W3 + C,
  data = dat,
  mix1 = c("X1", "X2", "X3"),
  mix2 = c("W1", "W2", "W3"),
  B_values = c(10, 20, 40),
  interaction = TRUE,
  q = 4,
  MCsize = nrow(dat),
  seed = 13
)
```

For quantization choice:

```{r, eval = FALSE}
q_sensitivity(
  f = Y ~ X1 + X2 + X3 + W1 + W2 + W3 + C,
  data = dat,
  mix1 = c("X1", "X2", "X3"),
  mix2 = c("W1", "W2", "W3"),
  q_values = c(3, 4, 5),
  interaction = TRUE,
  B = 10,
  seed = 13
)
```

# How to interpret these checks

The post-fit checks answer different questions:

- `support()` helps you inspect the intervention grid behind the stored surface
- `diagnostics(..., type = "bootstrap")` tells you whether the repeated-fit
  uncertainty calculation behaved cleanly
- `adequacy()` tells you whether the MSM is a reasonable summary of the exact
  fit-time surface on the stored grid
- the sensitivity helpers tell you whether conclusions are stable to practical
  analysis choices such as `MCsize`, `B`, and `q`

These checks are most useful together. A clean coefficient table is easier to
defend when the intervention grid is interpretable, bootstrap behavior is
stable, adequacy is acceptable, and the main conclusions do not move much under
reasonable sensitivity changes.

# Related Articles

- [Applied Workflow for Two-Mixture Quantile g-Computation](qgcompmulti-workflow.html)
- [Effect Scales, Intervals, and Prediction Scales](qgcompmulti-effect-scales.html)
- [Multiple Imputation and Parallel Workflows](qgcompmulti-mi-parallel.html)
- [Clustered Data and Repeated Measures](qgcompmulti-clustered-data.html)
