---
title: "Effect Scales, Intervals, and Prediction Scales"
description: >
  Understand how qgcomp.multi reports additive and ratio-scale MSM summaries,
  how bootstrap intervals are displayed, and why predictions stay on the
  response scale.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Effect Scales, Intervals, and Prediction Scales}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

# Introduction

Version `0.5.0` adds explicit estimand scales for the marginal structural model
(MSM) coefficients. This matters most for binary and count outcomes. A binary
outcome can be summarized on either an odds-ratio or risk-difference scale. A
count outcome can be summarized on either a rate-ratio or mean-difference scale.

The package now separates the following three ideas:

- the fitting scale used for coefficients, standard errors, covariance matrices,
  pooling, and confidence regions
- the display scale used in printed summaries, confidence intervals, and tidy
  output
- the response scale used for predicted risks, expected counts, and fitted
  surface plots
  
Those scales are related, but they are not interchangeable.

# Default estimand scales

The defaults follow the outcome family and link:

| Outcome family | Default estimand | Additive alternative |
| --- | --- | --- |
| `gaussian()` | `"mean_difference"` | none |
| `binomial(link = "logit")` | `"odds_ratio"` | `"risk_difference"` |
| `poisson(link = "log")` | `"rate_ratio"` | `"mean_difference"` |

For binomial and Poisson fits with other links, ratio-scale MSM fitting is not
available. The package falls back to additive response-scale summaries.

You can always make the choice explicit:

```{r, eval = FALSE}
fit_or <- qgcomp.glm.multi(
  f = Y ~ X1 + X2 + X3 + W1 + W2 + W3 + C,
  data = dat_binary,
  mix1 = c("X1", "X2", "X3"),
  mix2 = c("W1", "W2", "W3"),
  family = binomial(),
  estimand_scale = "odds_ratio",
  B = 100,
  seed = 13
)

fit_rd <- qgcomp.glm.multi(
  f = Y ~ X1 + X2 + X3 + W1 + W2 + W3 + C,
  data = dat_binary,
  mix1 = c("X1", "X2", "X3"),
  mix2 = c("W1", "W2", "W3"),
  family = binomial(),
  estimand_scale = "risk_difference",
  B = 100,
  seed = 13
)
```

# Fitting scale versus display scale

For additive estimands, the fitting and display scales are the same. For odds
ratios and rate ratios, they differ.

A ratio-scale MSM is fit on a log coefficient scale. That is the scale where
standard errors, covariance matrices, Rubin pooling, and confidence regions make
statistical sense. The package therefore keeps core machine-readable quantities
on the fitting scale:

```{r, eval = FALSE}
coef(fit_or)
vcov(fit_or)
tidy(fit_or)
```

For reporting, the same fitted object also gives display-scale values:

```{r, eval = FALSE}
fit_or
summary(fit_or)
confint(fit_or)
tidy(fit_or, conf.int = TRUE)
```

In a ratio-scale fit, the printed coefficient table includes both the
exponentiated ratio and the fitting-scale coefficient. That is deliberate. The
ratio is easier to read, but the fitting-scale coefficient is the one that has a
standard error with the usual large-sample interpretation.

# Binary outcomes

For binary outcomes fit with `family = binomial()`, the default v0.5.0 estimand
is an odds ratio. The data generator accepts ratio parameters on their natural
scale:

```{r, eval = FALSE}
dat_binary <- sim_mixture_data(
  n = 600,
  pA = 3,
  pB = 3,
  rho_within_A = 0.3,
  rho_within_B = 0.3,
  rho_between = 0.2,
  psi1 = 1.4,
  psi2 = 1.2,
  psi12 = 1.1,
  family = binomial(),
  baseline_risk = 0.2,
  seed = 123
)
```

Then fit the odds-ratio MSM:

```{r, eval = FALSE}
fit_or <- qgcomp.glm.multi(
  f = Y ~ X1 + X2 + X3 + W1 + W2 + W3 + C,
  data = dat_binary,
  mix1 = c("X1", "X2", "X3"),
  mix2 = c("W1", "W2", "W3"),
  family = binomial(),
  B = 100,
  seed = 13
)
summary(fit_or)
```

If the scientific question is additive risk, request that scale directly:

```{r, eval = FALSE}
fit_rd <- qgcomp.glm.multi(
  f = Y ~ X1 + X2 + X3 + W1 + W2 + W3 + C,
  data = dat_binary,
  mix1 = c("X1", "X2", "X3"),
  mix2 = c("W1", "W2", "W3"),
  family = binomial(),
  estimand_scale = "risk_difference",
  B = 100,
  seed = 13
)
```

Risk-difference fits stay on the identity scale. Their coefficients and display
values are therefore the same scale.

# Count outcomes

For Poisson-log fits, the default v0.5.0 estimand is a rate ratio:

```{r, eval = FALSE}
dat_count <- sim_mixture_data(
  n = 600,
  pA = 3,
  pB = 3,
  rho_within_A = 0.3,
  rho_within_B = 0.3,
  rho_between = 0.2,
  psi1 = 1.3,
  psi2 = 1.15,
  psi12 = 1.05,
  family = poisson(),
  baseline_rate = 1,
  seed = 123
)

fit_rr <- qgcomp.glm.multi(
  f = Y ~ X1 + X2 + X3 + W1 + W2 + W3 + C,
  data = dat_count,
  mix1 = c("X1", "X2", "X3"),
  mix2 = c("W1", "W2", "W3"),
  family = poisson(),
  B = 100,
  seed = 13
)

summary(fit_rr)
```

If expected-count differences are the target, use `estimand_scale =
"mean_difference"`.

# Confidence intervals

For single-fit objects, `confint()` supports three methods:

- `"wald"`, based on the coefficient estimate and stored bootstrap covariance
- `"percentile"`, based on empirical bootstrap quantiles
- `"basic"`, the reverse-percentile bootstrap interval

All three methods are computed on the MSM fitting scale. For odds-ratio and
rate-ratio estimands, the returned limits are exponentiated for display:

```{r, eval = FALSE}
confint(fit_or, method = "wald")
confint(fit_or, method = "percentile")
confint(fit_or, method = "basic")
```

Pooled multiple-imputation objects remain Wald-only. The pooled coefficients
and covariance are Rubin-pooled on the fitting scale, and then display columns
can be exponentiated for ratio summaries.

# Predictions and plots stay on the response scale

`predict()` and `plot()` answer a different question from `coef()`. They show
predicted risks, expected counts, or means under intervention regimes. Those
are response-scale quantities:

```{r, eval = FALSE}
predict(fit_or)
plot(fit_or)
```

This is true even when `fit_or` reports odds-ratio coefficients. A surface of
predicted risks is usually what users need to understand the fitted intervention
response.

For a direct comparison of two regimes on the fitted estimand scale, request an
MSM contrast:

```{r, eval = FALSE}
predict(
  fit_or,
  type = "msm_contrast",
  from = c(psi1 = 0, psi2 = 0),
  to = c(psi1 = 3, psi2 = 3),
  contrast_scale = "estimand",
  interval = TRUE,
  method = "basic"
)
```

For an odds-ratio fit, that contrast is reported as an odds ratio. For a
rate-ratio fit, it is reported as a rate ratio.

# Confidence regions

`confregion()` builds confidence regions for selected MSM coefficients in
single-fit objects:

```{r, eval = FALSE}
region <- confregion(fit_or, parm = c("psi1", "psi2"))
region
plot(region)
```

The region is based on the bootstrap covariance matrix and a chi-squared cutoff.
Its geometry is always on the fitting coefficient scale. For odds-ratio and
rate-ratio fits, that means the log coefficient scale. The package does not draw
nonlinear exponentiated ratio-scale ellipses in v0.5.0.

Pooled MI confidence regions are not supported in this release.

# Practical notes

A few rules prevent most interpretation mistakes:

- Use `coef()` and `vcov()` when downstream software needs fitting-scale
  coefficients and covariance matrices.
- Use print, `summary()`, `confint()`, and display columns from `tidy()` when
  writing ratio summaries for applied audiences.
- Use `predict()` and `plot()` when the question is about predicted risks,
  expected counts, or means.
- Use `predict(..., type = "msm_contrast", contrast_scale = "estimand")` when
  the question is a specific regime comparison on the odds-ratio or rate-ratio
  scale.
- Use `confregion()` for single-fit coefficient regions, and remember that
  those regions live on the fitting scale.
  
# Related articles
- [Applied Workflow for Two-Mixture Quantile g-Computation](qgcompmulti-workflow.html)
- [Multiple Imputation and Parallel Workflows](qgcompmulti-mi-parallel.html)
- [Clustered Data and Repeated Measures](qgcompmulti-clustered-data.html)
- [Diagnostics and Sensitivity Checks](qgcompmulti-diagnostics-sensitivity.html)
